---
title: "Building a Life Table"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(data.table)
library(lifeTableProcessing)

```

In this article we will walk through the steps of creating a life table from
input deaths and population data, including how to extend a life table to a
higher open age interval using an iterative method.

# Loading data

We begin by loading data for a single location-year. As an example, we will use
USA males in 2000 in 5-year age groups from HMD, initially collapsed to a lower
open age interval to show the effects of extension later

```{r}
lt_id_cols <- c("hmd_loc_id", "year", "source_type", "sex")
id_cols <- c(lt_id_cols, "age_start", "age_end")

open_age <- 110

lt <- collapse_open_age(
  lifeTableProcessing::hmd_vr_pop_usa_males_2000, 
  open_age = 70
)

```

We will also load regression parameters created from HMD life tables so that we
can create an initial set of life table parameters at higher ages to iterate

```{r}
coef_qx_diff <- lifeTableProcessing::hmd_extension_qx_diff
coef_ax_mx <- lifeTableProcessing::hmd_extension_ax_mx
coef_ax_qx <- lifeTableProcessing::hmd_extension_ax_qx
```

# Inital Life Table

From deaths and population, we can create an initial life table using naive
assumptions about $a_x$:

```{r}
lt[, mx := deaths / population]

lt[, ax := (age_end - age_start) / 2]
demCore::gen_u5_ax_from_mx(lt, id_cols)
lt[is.infinite(age_end), ax := demCore::mx_to_ax(mx, age_end - age_start)]

lt[, qx := demCore::mx_ax_to_qx(mx, ax, age_end - age_start)]
demCore::gen_lx_from_qx(lt, id_cols)
demCore::gen_dx_from_lx(lt, id_cols)
```

```{r, echo=FALSE}
knitr::kable(lt)
```


Then we can achieve better estimates of $a_x$ by recalculating $a_x$ from $d_x$,
recalculating $d_x$, and iterating until agreement is reached:

```{r}
lt_ax_prep <- demCore::iterate_ax(lt, id_cols)
```

```{r, echo=FALSE}
knitr::kable(lt_ax_prep)
```

# Extending the Life Table

Now we focus on extending the life table to the desired open age. We begin by
creating an initial set of $q_x$ values for ages between the input and desired
open age using regression parameters based on the difference in 
logit-transformed $q_x$ values

```{r}
lt_extended <- extend_qx_diff(
  lt_ax_prep[!is.infinite(age_end), c(..id_cols, "mx", "ax", "qx")],
  coef_qx_diff,
  lt_id_cols
)

lt_extended[
  is.na(mx) & age_start >= 5,
  mx := demCore::qx_ax_to_mx(qx, ax, age_end - age_start)
]

demCore::gen_lx_from_qx(lt_extended, id_cols)
demCore::gen_dx_from_lx(lt_extended, id_cols, assert_na = FALSE)
lt_extended[is.na(dx), dx := lx * qx]
```

```{r, echo=FALSE}
knitr::kable(lt_extended)
```


# Iterate to an optimal extension

With an initial projection of $q_x$, we can now iteratively solve for the ideal
extension by calculating life expectancy from the extended ages and comparing
it to the life expectancy at the original open age interval

```{r}
lt_extended[
  lt_ax_prep[is.infinite(age_end)],
  `:=`(
    terminal_age_start = i.age_start,
    mx_term = i.mx
  ),
  on = lt_id_cols
]

lt_extended[, age_length := age_end - age_start]
lt_extended[, age_end := NULL]
setnames(lt_extended, "age_start", "age")
setnames(coef_ax_qx, "age_start", "age")

iteration_results <- iterate_qx(
  dt = lt_extended,
  ax_params = coef_ax_qx[, -"age_end"],
  id_vars = c(lt_id_cols, "age"),
  n_iterations = 50
)

lt_iterated <- copy(iteration_results[[1]])
iteration_info <- copy(iteration_results[[2]])

setnames(lt_iterated, "age", "age_start")
lt_iterated[, age_end := shift(age_start, n = 1, type = "lead")]
lt_iterated[age_start == max(age_start), age_end := open_age]
setcolorder(lt_iterated, id_cols)
```

```{r, echo=FALSE}
knitr::kable(lt_iterated)
```

# Add New Open Age Interval

Finally, all that remains is to calculate the life table parameters for the
new open age interval

```{r}
lt_final <- rbind(
  lt_iterated,
  lt_iterated[1, ..lt_id_cols][, `:=`(age_start = open_age, age_end = Inf, qx = 1)],
  fill = TRUE
)

demCore::gen_lx_from_qx(lt_final, id_cols)
demCore::gen_dx_from_lx(lt_final, id_cols)
```

```{r, echo=FALSE}
knitr::kable(lt_final)
```

