---
title: "Extending Life Tables with HMD Regression Parameters"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

```{r setup}
library(arrow)
library(data.table)
library(lme4)
library(lifeTableProcessing)

cfg <- yaml::read_yaml("~/lifeTableProcessing.yml")
```

Some methods of extending a life table to higher open age intervals (such as 
iterative methods) require an initial set of life table parameters at these
older ages. One method of getting this information is by applying regression
parameters calculated from a known quality set of life tables with data out to
the desired open age already available. Here, we use life tables from HMD
to estimate a set of regression parameters that can be used to extend $q_x$ and 
$a_x$ to higher age groups.

# Preparing the Data

First, we must load HMD life table data

```{r}
id_vars <- c("hmd_loc_id", "year", "sex")

hmd_lts <- read_parquet(
  fs::path(cfg$dir_proj, "data/prepped/hmd_lifetables_5x1.parquet"),
  col_select = c(all_of(id_vars), "age_start", "age_end", "mx", "ax", "qx")
)

```

Then we can subset to the life tables we want to include in this analysis,
dropping location-years that should not be used to inform others

```{r}
hmd_lts <- hmd_lts[
  year >= 1950 &
    !(hmd_loc_id == "TWN" & year < 1980) &
    !(hmd_loc_id == "EST" & year > 2000) &
    !(hmd_loc_id %in% c("UKR", "BLR", "BGR") & year < 1970) &
    !(hmd_loc_id %in% c("LVA", "LTV") & year %between% c(1958, 1970)) &
    !(hmd_loc_id == "IRL" & year %between% c(1949, 1986)) &
    !(hmd_loc_id == "PRT" & year %between% c(1939, 1971)) &
    !(hmd_loc_id == "RUS" & year %between% c(1958, 1970)) &
    !(hmd_loc_id == "SVK" & year %between% c(1949, 1962)) &
    !(hmd_loc_id == "ESP" & year %between% c(1907, 1961))
]

```

Finally, we create computed variable for use in the different regressions

```{r}
hmd_lts[, mx_sq := mx^2]
hmd_lts[, qx_sq := qx^2]

hmd_lts[, qx_logit := demUtils::logit(qx)]

hmd_lts[
  j = `:=`(
    qx_logit_diff_to = qx_logit - shift(qx_logit, type = "lag"),
    qx_logit_diff_from = shift(qx_logit, type = "lead") - qx_logit
  ),
  by = id_vars
]

```


# Regressions

## Extend $q_x$ Using Logit Differences

First we get the $\text{logit}(q_x)$ differences from the prior and to the next
age group for age groups starting between 65 and 100:

```{r}
model_formula <- qx_logit_diff_from ~ qx_logit_diff_to + (1 | hmd_loc_id)

hmd_extension_qx_diff <- hmd_lts[
  age_start %between% c(65, 100) &
    !is.na(qx_logit_diff_to) & 
    !is.na(qx_logit_diff_from),
  j = .(fit = list(lme4::lmer(formula = model_formula, data = .SD))),
  by = .(sex, age_start, age_end)
]

hmd_extension_qx_diff[, `:=`(
  intercept = unlist(lapply(fit, \(f) fixef(f)[["(Intercept)"]])),
  slope = unlist(lapply(fit, \(f) fixef(f)[["qx_logit_diff_to"]])),
  fit = NULL
)]

```

```{r, eval=TRUE, echo=FALSE}
knitr::kable(lifeTableProcessing::hmd_extension_qx_diff)
```


## Extend $a_x$ using $m_x$

Here we fit a model:

$$
a_x \sim 1 + m_x + (m_x)^2
$$

for age groups starting between 65 and 105:

```{r}
hmd_extension_ax_mx <- hmd_lts[
  age_start %between% c(65, 105),
  j = .(fit = list(lm(ax ~ mx + mx_sq, data = .SD))),
  by = .(sex, age_start, age_end)
]

hmd_extension_ax_mx[, `:=`(
  par_con = unlist(lapply(fit, \(f) coef(f)[["(Intercept)"]])),
  par_mx = unlist(lapply(fit, \(f) coef(f)[["mx"]])),
  par_smx = unlist(lapply(fit, \(f) coef(f)[["mx_sq"]])),
  fit = NULL
)]

```

```{r, eval=TRUE, echo=FALSE}
knitr::kable(lifeTableProcessing::hmd_extension_ax_mx)
```


## Extend $a_x$ using $q_x$

Here we fit a model:

$$
a_x \sim 1 + q_x + (q_x)^2
$$

for age groups starting between 65 and 105:

```{r}
hmd_extension_ax_qx <- hmd_lts[
  age_start %between% c(65, 105),
  j = .(fit = list(lm(ax ~ qx + qx_sq, data = .SD))),
  by = .(sex, age_start, age_end)
]

hmd_extension_ax_qx[, `:=`(
  par_con = unlist(lapply(fit, \(f) coef(f)[["(Intercept)"]])),
  par_qx = unlist(lapply(fit, \(f) coef(f)[["qx"]])),
  par_sqx = unlist(lapply(fit, \(f) coef(f)[["qx_sq"]])),
  fit = NULL
)]

```

```{r, eval=TRUE, echo=FALSE}
knitr::kable(lifeTableProcessing::hmd_extension_ax_qx)
```


```{r, include=FALSE}

# Run this if you want to update the data included in the package

usethis::use_data(
  hmd_extension_qx_diff,
  hmd_extension_ax_mx,
  hmd_extension_ax_qx,
  overwrite = TRUE
)

```

