---
title: "Preparing HMD Data"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

```{r setup}
library(data.table)
library(lifeTableProcessing)

cfg <- yaml::read_yaml("~/lifeTableProcessing.yml")
```

# Loading HMD Data

We will be working with HMD data by statistic (meaning all countries) located at
https://mortality.org/Data/ZippedDataFiles. You will need to register an account
with HMD before being able to access this data. Specifically, we need:

* Period death counts
* Period population estimates
* Period Life tables (female and male)

We need the **5x1** data, meaning "five year age group by single year time
interval". After extracting the data, you should have a set of directories like:

```
data/raw
├── Deaths_5x1
├── fltper_5x1
├── mltper_5x1
└── Population5
```

where each directory contains files for each location in HMD.

To load the data, run:

```{r}
deaths_5x1 <- prep_hmd_deaths_5x1(fs::path(cfg$dir_proj, "data/raw/Deaths_5x1"))

pop_5x1 <- prep_hmd_pop_5x1(fs::path(cfg$dir_proj, "data/raw/Population5"))

lt_5x1 <- prep_hmd_period_lt_5x1(
  dir_input_male = fs::path(cfg$dir_proj, "data/raw/mltper_5x1"),
  dir_input_female = fs::path(cfg$dir_proj, "data/raw/fltper_5x1")
)

```

# Preparing the Data

In order to use the deaths and population data in _lifeTableProcessing_, we
need to apply some additional formatting:

```{r}
dt_hmd <- merge(
  deaths_5x1,
  pop_5x1,
  by = c("hmd_loc_id", "year", "sex", "age_start", "age_end"),
  all = TRUE
)

dt_hmd[, data_origin := "hmd_5x1"]
dt_hmd[, source_type := "VR"]

setcolorder(dt_hmd, c("data_origin", "source_type"))

```

We also want to only include data starting from 1950, exclude any
location-years with boundary changes, and exclude and incomplete data. With
this final subset of data, we can calculate a mortality rate:

```{r}
dt_hmd_sub <- dt_hmd[
  year >= 1950 &
    is.na(territory_change_status) &
    !is.na(deaths) &
    !is.na(population),
  -c("territory_change_status")
]

dt_hmd_sub[, mx := deaths / population]
```

# Saving the results

Prepared data can be saved in whatever way is convenient, but other examples
in this package assume the deaths and population data is saved as an 
[Arrow Dataset](https://arrow.apache.org/docs/r/articles/dataset.html).

```{r}
arrow::write_dataset(
  dataset = dt_hmd_sub,
  path = fs::path(cfg$dir_proj, "data/prepped/input_vr_pop"),
  format = "parquet",
  partitioning = "data_origin"
)

arrow::write_parquet(
  lt_5x1,
  fs::path(cfg$dir_proj, "data/prepped/hmd_lifetables_5x1.parquet")
)

```

